Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.

### 2024-03-11 apples_and_oranges.qmd
Extracting coefficients for meta-analyses

We ran a random effects meta-analysis in R's Meta package to compare the active versus control arms of psychotherapy trials on key variables of interest regarding the nature and intensity of the interventions. We extracted data pertaining to the number, duration and intensity of sessions, and the total cumulative hours and period of the intervention. Where a range was provided, the maximum was encoded (e.g. if a paper reported that an intervention involved 8-10 sessions lasting 50-60 minutes, we encoded the number and duration of sessions as 10 and 60, respectively). If sessions varied in frequency across an intervention, we calculated an average by dividing total number of sessions by length of intervention period. Similarly, if the length of sessions varied across the course of the intervention, we calculated a weighted average. Phone call, web-chat and online sessions were encoded as sessions, however guided self-help components were not.

@@

Psychotherapy trials featured a higher percentage of women when compared to medication trials. This subgroup difference in gender composition reached statistical significance (stats). Excluding sub-clinical studies yielded very similar results (stats). This difference in gender composition however did not reach statistical significance when only CBT, fluoxetine and escitalopram trials were included (stats).

@@

# Define function for baseline metaanalyses  ----------------------------------------------

# subgroup_df <- data.frame(psy_or_med = c(0, 1))
# 
# result <- list()
# name_vec <- character()
# 
# # Loop over rows in the subgroup_df
# for (i in 1:nrow(subgroup_df)) {
#   
#   # Call the metagen function for each row
#   result[[i]] <- metagen(
#     TE = percent_women,
#     seTE = percent_women_std_error,
#     studlab = new_study_id,
#     data = df_for_gender_metan[df_for_gender_metan$psy_or_med == subgroup_df$psy_or_med[i], ],
#     sm = "",
#     fixed = FALSE,
#     random = TRUE,
#     method.tau = "REML",
#     hakn = TRUE,
#     title = paste("Subgroup:", subgroup_df$psy_or_med[i])
#   )
#   
#   name_vec[i] = paste("Subgroup:", subgroup_df$psy_or_med[i])
# }
# 
# # Now result contains the metagen results for each subgroup
# # name_vec contains names for each subgroup
# 
# ## Function to extrac the key results for graphing and tabulating.
# 
# extract_stats <- function(result) {
#  
#   # get the summary table with results
#   statistics <- summary(result)
#   
#   # Extract specific stats
#   extracted_stats <- c(
#     k_study = statistics$k.study,
#     TE_random = statistics$TE.random,
#     lower_random = statistics$lower.random,
#     upper_random = statistics$upper.random,
#     pval = statistics$pval.random,
#     tau2 = statistics$tau2,
#     lower_tau2 = statistics$lower.tau2,
#     upper_tau2 = statistics$upper.tau2
#   )
#   
#   return(extracted_stats)
# }
# 
# extract_stats(result[[1]])
# extract_stats(result[[2]])
# 
# # create a dataframe with the results
# statistics_indiv_metan <- lapply(result, extract_stats) 
# statistics_indiv_metan <- data.frame(do.call("rbind", statistics_indiv_metan))
# statistics_indiv_metan$condition <- name_vec

@@

met_perc_women <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

# met_perc_women_overall <- metagen(TE = percent_women,
#                            seTE = percent_women_std_error,
#                            studlab = new_study_id,
#                            data = df_for_gender_metan,
#                            sm = "",
#                            fixed = FALSE,
#                            random = TRUE,
#                            method.tau = "REML",
#                            hakn = TRUE,
#                            title = "percentage women across studies")
# 
# subgroup_metan <-update.meta(met_perc_women_overall  , 
#                               subgroup = df_for_gender_metan$psy_or_med, subgroup.name = "treatment modality",
#                               tau.common = FALSE)

# Use the summary function with the 'byvar' argument
summary_met_perc_women <- summary(met_perc_women, byvar = "psy_or_med")

# Extract specific statistics for each subgroup
subgroup_names <- summary_met_perc_women$subgroup.name
TE_random <- summary_met_perc_women$TE.random
seTE_random <- summary_met_perc_women$seTE.random
lower_random <- summary_met_perc_women$lower.random
upper_random <- summary_met_perc_women$upper.random
pval <- summary_met_perc_women$pval.random
tau2 <- summary_met_perc_women$tau2
lower_tau2 <- summary_met_perc_women$lower.tau2
upper_tau2 <- summary_met_perc_women$upper.tau2

# Create a data frame to store the results
result_df <- data.frame(
  subgroup = subgroup_names,
  TE_random = TE_random,
  seTE_random = seTE_random,
  lower_random = lower_random,
  upper_random = upper_random,
  pval = pval,
  tau2 = tau2,
  lower_tau2 = lower_tau2,
  upper_tau2 = upper_tau2
)

knitr:: kable(result_df, "latex", caption = "Gender composition")

@@

met_perc_women_clin <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan_clin,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across clinical studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

@@

df_for_gender_metan_cbt <- df_for_gender_metan %>%
  filter(new_study_id %in% cbt_meds_study_ids) %>% 
  filter(!(arm_effect_size == "cohens_d_active" & !(type %in% c("Fluoxetine", "Escitalopram", "cbt"))))

met_perc_women_cbt <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across cbt and med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)
```

@@

Description of results X + table.

@@

@@

@@

df_for_age_metan_cbt <- df_for_age_metan %>%
  filter(new_study_id %in% cbt_meds_study_ids) %>% 
  filter(!(arm_effect_size == "cohens_d_active" & !(type %in% c("Fluoxetine", "Escitalopram", "cbt"))))

met_age_cbt <- metagen(TE = mean_age,
                           seTE = se_age,
                           studlab = new_study_id,
                           data = df_for_age_metan_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "age across cbt + med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)
```
### Baseline severity

Describe results here

@@

@@

@@

df_for_baseline_severity_cbt <- df_for_baseline_severity %>%
  filter(new_study_id %in% cbt_meds_study_ids) %>% 
  filter(!(arm_effect_size == "cohens_d_active" & !(type %in% c("Fluoxetine", "Escitalopram", "cbt"))))

met_sev_cbt <- metagen(TE = perc_baseline_overall,
                           seTE = perc_se_baseline,
                           studlab = new_study_id,
                           data = df_for_baseline_severity_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "baseline severity cbt + med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)
```

>>>>

We ran t-tests to compare the active versus control arms of psychotherapy trials on key variables of interest regarding the nature and intensity of the interventions. We extracted data pertaining to the number, duration and intensity of sessions, and the total cumulative hours and period of the intervention. Where a range was provided, the maximum was encoded (e.g. if a paper reported that an intervention involved 8-10 sessions lasting 50-60 minutes, we encoded the number and duration of sessions as 10 and 60, respectively). If sessions varied in frequency across an intervention, we calculated an average by dividing total number of sessions by length of intervention period. Similarly, if the length of sessions varied across the course of the intervention, we calculated a weighted average. Phone call, web-chat and online sessions were encoded as sessions, however guided self-help components were not.

@@

Psychotherapy trials featured a significantly higher percentage of women when compared to medication trials (`results_met_perc_women_overall$pval_between`). On average, samples were comprised of `r results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 1]`% women across psychotherapy trials and `r results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 0]`% women across medication trials.  Excluding sub-clinical studies yielded very similar results (p = `results_met_perc_women_clin$pval_between`). This difference in gender composition however did not reach statistical significance when only CBT, fluoxetine and escitalopram trials were included (p = `results_met_perc_women_cbt$pval_between`).

@@

# This is the function to extract statistics

extract_stats <- function(df) {

  # get the summary table with results
  statistics <- summary(df, byvar = "psy_or_med")

  df_stats <- data.frame(
  subgroup = statistics$subgroup.levels,
  k_study = statistics$k.study.w,
  TE_random = statistics$TE.random.w,
  seTE_random = statistics$seTE.random.w,
  lower_random = statistics$lower.random.w,
  upper_random = statistics$upper.random.w,
  tau2 = statistics$tau2.w)

  # Add an empty row at the top and column on right (this is for the test for subgroup differences p value)
  df_stats <- rbind(data_frame(subgroup = NA, k_study = NA, TE_random = NA, seTE_random = NA, 
                             lower_random = NA, upper_random = NA, tau2 = NA), df_stats)
  df_stats <- cbind(df_stats, pval_between = NA)

  # Set the value in the last row and last column as the between groups p val
  df_stats[1, "pval_between"] <- statistics$pval.Q.b.random

  return(df_stats)

@@

met_perc_women_overall <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across studies")

met_perc_women_overall <- update(met_perc_women_overall,
                              subgroup = df_for_gender_metan$psy_or_med, subgroup.name = "treatment modality",
                              tau.common = FALSE)

# Apply the function to the meta-analysis result
results_met_perc_women_overall <- extract_stats(met_perc_women_overall)

knitr:: kable(results_met_perc_women_overall, caption = "Gender composition")

@@

met_perc_women_clin <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan_clin,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across clinical studies")

met_perc_women_clin <- update(met_perc_women_clin,
                           subgroup = df_for_gender_metan_clin$psy_or_med, subgroup.name = "treatment modality",
                           tau.common = FALSE)

# Apply the function to the meta-analysis result
results_met_perc_women_clin <- extract_stats(met_perc_women_clin)

knitr:: kable(results_met_perc_women_clin, caption = "Gender composition in clinical studies")

@@

df_for_gender_metan_cbt <- df_for_gender_metan %>%
  filter(new_study_id %in% cbt_meds_study_ids)

met_perc_women_cbt <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across cbt and med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

# Apply the function to the meta-analysis result
results_met_perc_women_cbt <- extract_stats(met_perc_women_cbt)

knitr:: kable(results_met_perc_women_cbt, caption = "Gender composition in studies with CBT, escitalopram or fluoxetine")

@@

Mean age was `r results_met_age$TE_random[results_met_perc_women_overall$subgroup == 1]` (`r results_met_age$seTE_random[results_met_perc_women_overall$subgroup == 1]`) across psychotherapy trials and `r results_met_age$TE_random[results_met_perc_women_overall$subgroup == 0]` (`r results_met_age$seTE_random[results_met_perc_women_overall$subgroup == 0]`) across medication trials, with no significant between group differences (p = `r results_met_age$pval_between`). There were no significant differences in mean age between modalities on further sensitivity analyses.

@@

results_met_age <- extract_stats(met_age)

knitr:: kable(results_met_age, caption = "Average age in psychotherapy vs medication trials")

@@

results_met_age_clin <- extract_stats(met_age_clin)

knitr:: kable(results_met_age_clin, caption = "Average age in clinical studies")

@@

df_for_age_metan_cbt <- df_for_age_metan %>%
  filter(new_study_id %in% cbt_meds_study_ids)

met_age_cbt <- metagen(TE = mean_age,
                           seTE = se_age,
                           studlab = new_study_id,
                           data = df_for_age_metan_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "age across cbt + med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

results_met_age_cbt <- extract_stats(met_age_cbt)

knitr:: kable(results_met_age_cbt, caption = "Average age in cbt, fluox, escit studies")

```
### Baseline severity

On average, severity of depressive symptoms at baseline was significantly higher in medication trials when compared to psychotherapy trials, for both the overall sample (p = `results_met_sev$pval_between`) and when taking only those studies where the active arm was eiter CBT, fluoxetine or escitalopram (p = `results_met_sev_cbt$pval_between`). When excluding psychotherapy studies that recruited samples with sub-clinical depression, this difference no longer reached statistical significance (p = `results_met_sev_clin$pval_between`).

@@

results_met_sev <- extract_stats(met_sev)

knitr:: kable(results_met_sev, caption = "Baseline Depression Severity in Psychotherapy vs Medication trials")

@@

results_met_sev_clin <- extract_stats(met_sev_clin)

knitr:: kable(results_met_sev_clin, caption = "Baseline Depression Severity in Psychotherapy vs Medication trials, excluding subclinical studies")

@@

df_for_baseline_severity_cbt <- df_for_baseline_severity %>%
  filter(new_study_id %in% cbt_meds_study_ids) %>%

met_sev_cbt <- metagen(TE = perc_baseline_overall,
                           seTE = perc_se_baseline,
                           studlab = new_study_id,
                           data = df_for_baseline_severity_cbt,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "baseline severity cbt + med studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

results_met_sev_cbt <- extract_stats(met_sev_cbt)

knitr:: kable(results_met_sev_cbt, caption = "Baseline Depression Severity in Psychotherapy vs Medication trials with CBT, fluoxetine and escitaloptram as the active intervention")

ggplot(df_for_baseline_severity, aes(x = factor(psy_or_med), y = perc_baseline_mean, color = factor(psy_or_med))) +
  geom_point(position = position_jitter(width = 0.2), size = 2) +
  labs(
    title = "Baseline Depression Severity in Psychotherapy vs Medication trials",
    x = "Treatment Modality",
    y = "Baseline Depression Levels (transformed)",
    color = "Treatment Type"
  ) +
  scale_color_discrete(labels = c("Medication", "Psychotherapy"))
